Linear vertical shear instability in astrophysical disks: a numerical study

### 2014-12-04 linear_iso.tex
new arguments on growth rate limit

With these approximations we can re-write the governing equation as
\begin{align}\label{iso_ode1}
  &\frac{d}{dz}\left(\rho e^{-\ii k_x r
      \Omega^2/\Omega_k^2}\frac{dW}{dz}\right) +
  \sigma^2\left(\frac{1}{c_s^2}+\frac{k_x^2}{\Omega_k^2}\right) \rho W e^{-\ii k_x r
    \Omega^2/\Omega_k^2} \notag\\
  &=0.
\end{align}
Next, we multiply Eq. \ref{iso_ode1} by $W^{*}\exp{\left(\ii k_x r
    \Omega^2/\Omega_k\right)}$, integrate 
vertically and assume boundary terms vanish to obtain
\begin{align}\label{neccessary_cond}
   &\sigma^2\left(\frac{1}{c_s^2}+\frac{k_x^2}{\Omega_k^2}\right)\int_{-\infty}^{\infty}\rho|W|^2dz \notag\\
  &=\int_{-\infty}^{\infty}\rho\left|\frac{dW}{dz}\right|^2dz 
  +\frac{\ii k_x r}{\Omega_k^2}\int_{-\infty}^{\infty}\rho\frac{d\Omega^2}{dz}W^*\frac{dW}{dz}dz. 
\end{align}
It follows that for instability ($\imag\sigma>0$), it is neccessary to
have $d\Omega^2/dz\neq 0$, i.e. vertical shear.

\subsection{Thin-disk limit}
The full expressions for the equilibrium density and vertical shear
depends on $z$ in a non-trivial manner. Thus, to simplify the problem 
we consider thin-disks ($\epsilon\ll1$), 
so we may expand the background density and vertical shear in powers
of $z/r$. To lowest order we obtain
\begin{align}
  &\rho(z) \simeq \rho_{00} \exp{\left(-\frac{z^2}{2\Hiso^2}\right)},\label{thin_dens}\\
  &\frac{d\Omega^2}{dz} \simeq \frac{q\Omega_k^2z}{r^2}. \label{thin_vshear}
\end{align}

With this approximation and those in \S\ref{integral_relation}, the
governing equation becomes
\begin{align}\label{iso_ode3}
  \frac{d^2W}{d\hat{z}^2} - \left(1 + \ii q\epsilon
    \hat{k}\right)\hat{z}\frac{dW}{d\hat{z}} +
  \left(\frac{\sigma}{\Omega_k}\right)^2\left(1+\hat{k}^2\right)W = 
  0, 
\end{align}
where we have introduced the dimensionless vertical co-ordinate
$\hat{z} \equiv z/\Hiso$ and wavenumber $\hat{k} \equiv k_x \Hiso$. We 
remark that for large $|\hat{k}|$, Eq. \ref{iso_ode3} is the same as
that derived by \cite{nelson13}, although we have taken a different
route.

In the thin-disk approximation the integral relation
Eq. \ref{neccessary_cond} becomes
\begin{align}\label{integral_relation_thin}
 & \left(\frac{\sigma}{\Omega_k}\right)^2\left(1+\hat{k}^2\right)
  \int_{-\infty}^{\infty}w|W|^2d\hat{z}  \notag\\ 
  & =\int_{-\infty}^{\infty}w\left|\frac{dW}{d\hat{z}}\right|^2d\hat{z}
  +\ii q \epsilon
  \hat{k}\int_{-\infty}^{\infty}w\hat{z}W^*\frac{dW}{d\hat{z}}d\hat{z}.    
\end{align}
where 
\begin{align}
  w(\hat{z}) \equiv \exp{\left(-\frac{\hat{z}^2}{2}\right)}.  
\end{align}

@@

Linearizing the integral relation in the thin-disk limit, Eq. \ref{integral_relation_thin}, and taking the
imaginary part, we have
\begin{align}
  \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
  \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
  \He_n^2(\hat{z}) d\hat{z} \notag\\
  = \delta q \epsilon \hat{k} 
  \int_{-\infty}^{\infty}
  w(\zhat)\hat{z}\He_n(\hat{z})\frac{d\He_n}{d\zhat} d\hat{z}
\end{align}
Recognizing $\hat{z} = \He_1$ and using $d\He_n/d\hat{z} = n
\He_{n-1}$ we have
 \begin{align}
   \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
   \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
   \He_n^2(\hat{z}) d\hat{z} \notag\\
   =\delta q \epsilon \hat{k} 
   \int_{-\infty}^{\infty}
   w(\zhat)\He_1(\hat{z})\He_n(\hat{z})n\He_{n-1}(\hat{z})d\hat{z}. 
 \end{align}

>>>>

With these approximations we can re-write the governing equation as
\begin{align}\label{iso_ode1}
  &\frac{d}{dz}\left(\rho e^{-\ii k_x r
      \Omega^2/\Omega_k^2}\frac{dW}{dz}\right) +
  \sigma^2\left(\frac{1}{c_s^2}+\frac{k_x^2}{\Omega_k^2}\right) \rho W e^{-\ii k_x r
    \Omega^2/\Omega_k^2} \notag\\
  &=0.
\end{align}
Next, we multiply Eq. \ref{iso_ode1} by $W^{*}\exp{\left(\ii k_x r
    \Omega^2/\Omega_k\right)}$ and integrate 
vertically from $z=-\zmax$ to $z=\zmax$. We assume either $W$ or
$dW/dz$ vanishes at $z=\pm \zmax$, or that $\zmax$ is sufficiently
large so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}\label{neccessary_cond}
   &\sigma^2\left(\frac{1}{c_s^2}+\frac{k_x^2}{\Omega_k^2}\right)\int_{-\zmax}^{\zmax}\rho|W|^2dz \notag\\
  &=\int_{-\zmax}^{\zmax}\rho\left|\frac{dW}{dz}\right|^2dz 
  +\frac{\ii k_x r}{\Omega_k^2}\int_{-\zmax}^{\zmax}\rho\frac{d\Omega^2}{dz}W^*\frac{dW}{dz}dz. 
\end{align}
It follows that for instability ($\imag\sigma>0$), it is neccessary to
have $d\Omega^2/dz\neq 0$, i.e. vertical shear.

\subsection{Maximum growth rate}   
Here we show that the growth rate is limited by the maximum vertical
shear in the domain. It is convenient to introduce the dimensionless
vertical co-ordinate $\hat{z} = z/\Hiso$ and wavenumber
$\hat{k}=k_x\Hiso$. Then 
the integral relation Eq. \ref{neccessary_cond} becomes 
\begin{align}\label{integral_relation1}
  &\left(\frac{\sigma}{\Omega_k}\right)^2\left(1+\hat{k}^2\right)
  \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho|W|^2d\hat{z}  
  \notag\\ 
  &=\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho\left|\frac{dW}{d\hat{z}}\right|^2d\hat{z}      
  +\ii
  \epsilon\hat{k}q\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
  f(\hat{z}) W^*\frac{dW}{d\hat{z}}d\hat{z},    
\end{align}   
where $f(\hat{z})$ is defined such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}

For convenience, we write $\hat{\sigma}\equiv\sigma/\Omega_k = \hat{\omega} +
\ii\hat{\nu}$ with $\hat{\omega},\hat{\nu}$ being the dimensionless
mode frequency and growth rates. Then the real and imaginary parts of
Eq. \ref{integral_relation1} are
\begin{align}
  &\left(\hat{\omega}^2-\hat{\nu}^2\right)\left(1+\hat{k}^2\right)
   \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho|W|^2d\hat{z} -
  \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho\left|\frac{dW}{d\hat{z}}\right|^2d\hat{z}
  \notag\\
  &=\real\left[\ii
  \epsilon\hat{k}q\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
 f(\hat{z}) W^*\frac{dW}{d\hat{z}}d\hat{z}\right], \notag\\
& 2\hat{\omega}\hat{\nu}\left(1+\hat{k}^2\right)
   \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho|W|^2d\hat{z}\notag\\
   &=\imag\left[\ii
  \epsilon\hat{k}q\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
 f(\hat{z}) W^*\frac{dW}{d\hat{z}}d\hat{z}\right].
\end{align}
Adding the square of these equations give
\begin{align}
&\left[|\hat{\sigma}|^2\left(1+\hat{k}^2\right)
   \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho|W|^2d\hat{z} -
  \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho\left|\frac{dW}{d\hat{z}}\right|^2d\hat{z}\right]^2\notag\\
&+4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
   |W|^2d\hat{z}\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho\left|\frac{dW}{d\hat{z}}\right|^2d\hat{z}\notag\\
   &=\left|\ii
  \epsilon\hat{k}q\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
 f(\hat{z}) W^*\frac{dW}{d\hat{z}}d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
&4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
   |W|^2d\hat{z}\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho\left|\frac{dW}{d\hat{z}}\right|^2d\hat{z}\notag\\
   &\leq\left|
  \epsilon\hat{k}q\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
 f(\hat{z}) W^*\frac{dW}{d\hat{z}}d\hat{z}\right|^2.
\end{align}
% We now take the same approach as in the previous section. We first
% assume an neutral solution corresponding to $q\equiv0$, for which
% $\sigma$ is real and we can take $W$ to be real as well. We then
% set $q\to\delta q$ and linearize. This procedure means
% \begin{align}
%   &\left(\frac{\sigma}{\Omega_k}\right)^2\left(1+\hat{k}^2\right)
%   \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho W^2d\hat{z}  
%   =\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho\left(\frac{dW}{d\hat{z}}\right)^2d\hat{z},\\ 
%   &\frac{2\sigma\imag{\delta\sigma}}{\Omega_k^2}\left(1+\hat{k}^2\right)
%   \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho W^2d\hat{z}\notag\\
%   &=\epsilon\hat{k}\delta q\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
%   f(\hat{z}) W\frac{dW}{d\hat{z}}d\hat{z}. 
% \end{align}
% Eliminating $\sigma^2$ between the above equations, we obtain
% \begin{align}\label{sigma_finite_domain}
%   &4
%   \left[\frac{\imag{\left(\delta\sigma\right)}}{\Omega_k}\right]^2\left(1+\hat{k}^2\right) 
%   \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
%   W^2d\hat{z}\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho\left(\frac{dW}{d\hat{z}}\right)^2d\hat{z}\notag\\
%   &=\left(\epsilon\hat{k}\delta q\right)^2\left(\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
%   f(\hat{z}) W\frac{dW}{d\hat{z}}d\hat{z}\right)^2. 
% \end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  &\left( \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
    |W|\left|\frac{dW}{d\hat{z}}\right|d\hat{z}\right)^2\notag\\&\leq
  \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho 
  |W|^2d\hat{z}\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho\left|\frac{dW}{d\hat{z}}\right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
    f(\hat{z}) W^*\frac{dW}{d\hat{z}}d\hat{z}\right|\leq \int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
  \left|f(\hat{z})W^*\frac{dW}{d\hat{z}}\right|d\hat{z} \notag\\
  \leq
  \mathrm{max}\left(|f|\right)\int_{-z_\mathrm{max}}^{z_\mathrm{max}}\rho
  |W|\left|\frac{dW}{d\hat{z}}\right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$z\in[-\zmax,\zmax]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  \hat{\nu}\equiv\left|\frac{\imag{(\sigma)}}{\Omega_k}\right| \leq
  \frac{\epsilon |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|). 
\end{align}
Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), it
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
value of the vertical shear in the domain considered.

For the vertical shear profile of the vertically isothermal
disk, $f(\hat{z}) =
\hat{z}\left(1+\epsilon^2\hat{z}^2\right)^{-3/2}$. If we consider
$\zmax\to\infty$,  then the growth rate is limited by the global
maximum of the vertical shear, which occurs at $\zhat
=1/\sqrt{2}\epsilon$ for the vertically isothermal disk.  
For vertical domains smaller than this height, the growth rate is
limited by $|f|$ at the vertical boundaries considered. For 
$|\epsilon\hat{z}|\ll1$ we have $f\simeq \hat{z}$, in which case we
expect growth rates to increase linearly with the vertical domain
size.

\subsection{Thin-disk limit}
We can make further progress for thin-disks ($\epsilon\ll1$), 
so we may expand the background density and vertical shear in powers
of $z/r$. To lowest order we obtain 
\begin{align}
  &\rho(\hat{z}) \simeq \rho_{00}
  \exp{\left(-\frac{\hat{z}^2}{2}\right)} \equiv \rho_{00}w(\hat{z}),\label{thin_dens}\\
  &\frac{d\Omega^2}{d\hat{z}} \simeq \epsilon^2q\Omega_k^2\hat{z}. \label{thin_vshear}
\end{align}
With this approximation and those in \S\ref{integral_relation}, the
governing equation becomes 
\begin{align}\label{iso_ode3}
  \frac{d^2W}{d\hat{z}^2} - \left(1 + \ii q\epsilon
    \hat{k}\right)\hat{z}\frac{dW}{d\hat{z}} +
  \left(\frac{\sigma}{\Omega_k}\right)^2\left(1+\hat{k}^2\right)W = 
  0.
\end{align}
We remark that for large $|\hat{k}|$, Eq. \ref{iso_ode3} is the same as
that derived by \cite{nelson13}, although we have taken a different
route.

To complete the problem we must specify boundary conditions. The
maximum vertical domain size we should consider is limited by the thin disk
approximation (i.e. $z\lesssim r$). It is most common to take $\zmax\to\infty$
and impose that the kinetic  energy remain bounded at infinity.  This
is appropriate for the density because both the true density profile
and its thin-disk approximation decay rapidly away from the midplane.

However, the thin-disk representation of vertical shear diverges with 
$\hat{z}$, unlike the true vertical shear profile which eventually
decays.  Thus taking $\zmax\to$ will permit unphysically large growth
rates (as demonstrated below). Nevertheless, the infinite disk is
useful to consider since it permits an analytical discussion.

@@

Linearizing the integral relation Eq. \ref{integral_relation1} in the
thin-disk limit, and taking the
imaginary part, we have
\begin{align}
  \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
  \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
  \He_n^2(\hat{z}) d\hat{z} \notag\\
  = \delta q \epsilon \hat{k} 
  \int_{-\infty}^{\infty}
  w(\zhat)\hat{z}\He_n(\hat{z})\frac{d\He_n}{d\zhat} d\hat{z}
\end{align}
Recognizing $\hat{z} = \He_1$ and using $d\He_n/d\hat{z} = n
\He_{n-1}$ we have
 \begin{align}
   \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
   \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
   \He_n^2(\hat{z}) d\hat{z} \notag\\
   =\delta q \epsilon \hat{k} 
   \int_{-\infty}^{\infty}
   w(\zhat)\He_1(\hat{z})\He_n(\hat{z})n\He_{n-1}(\hat{z})d\hat{z}. 
 \end{align}
